Anxiety: a decision-theoretic perspective

### 2019-02-03 manuscript/manuscript.tex
many methods updates

\author{
Samuel Zorowitz \\
Princeton Neuroscience Institute\\
Princeton University\\
Princeton, NJ 08540 \\
\texttt{zorowitz@princeton.edu} \\
\And
Ida Momennejad \\
columbia University\\
New York, NY 10027 \\
\texttt{ida.m@columbia.edu} \\
\And
Nathaniel Daw \\
Princeton Neuroscience Institute\\
Princeton University\\
Princeton, NJ 08540 \\
\texttt{ndaw@princeton.edu} \\
}

@@

\section{Introduction}
Many psychiatric disorders involve distortions in models of the external world
and, correspondingly, aberrations of learning and decision making. As a normative
framework, decision theory allows us to describe and pose quantita­tive questions
about optimal and approximately optimal choice behavior \citep{DayanDaw2008} and
is, therefore, a critical tool for modeling, understanding, and predicting the
psychological and neurobiological mechanisms responsible for the abnormalities
observed in psychiatric behavior \citep{HuysDawDayan2015}. Many recent investigations
employing this approach have provided important insights into psychiatric symptoms,
including anhedonia in major depression \citep{Rutledge2017}, mood instability
in bipolar disorder \citep{EldarNiv2015, EldarDolanNiv2016},  habitual decision
making in disorders of compulsivity \citep{gillan2016}, and hallucinations in
schizophrenia \citep{powers2017, corlett2018}.

Anxiety disorders (generalized anxiety disorder, social anxiety disorder, panic
disorder, obsessive-compulsive disorder) represent the most common family of
psychiatric disorders (citation). Despite the differences in the content of worry,
all anxiety disorders share two core features: pessimistic inference and excessive
avoidance. Pessimistic inference describes a tendency to appraise threatening
situations as more frequent and/or more severe than they objectively are. This has
been reliably observed experimentally, as in fear conditioning wherein individuals
with anxiety exhibit exaggerated fear towards safe cues (CS-) and extinguished
threatening cues \citep{lissek2005, Duits2015}, and clinically, through self-report
(pick 3 citaitons). Excessive avoidance describes the corresponding tendency for
individuals with anxiety to consistently flee or avoid threatening situations.
Avoidance is often chosen even at the expense of potential reward (approch-avoidance
conflict), and is frequently a major impediment to recovery \citep{Arnaudova2017}.

Recent studies have attempted to provide some computational models of this behavior
\citep{Moutoussis2008, Maia2010}. For example, \citep{Maia2010} was able to show
how reinforcement learning could explain sustained avoidance in the absence of
repeated reinforcement, observed in rodents \citep{servatius2008} and something
not explained by two-factor theory \citep{Krypotos2015}. As has been discussed
elsewhere \citep{Moutoussis2017}, such accounts only explain the maintenance of
avoidance in short timescales, but do not explain why resistant to change and
why they do not go away in general. Such theories also do not explain the magnitude
of pessimistic inference, or the fear generalization. A broader account is needed.

In this article, we put forward a decision theoretic explanation of the core
symptoms of anxiety. In sequential decision environments, we show how a loss of
optimistic inference results in pessimistic inference, chronic avoidance, and
generalization of fear.

@@

Here, we briefly review the formalism of Markov decision processes (MDPs), which
provide the foundation for our results. For more complete treatments of the
subject, see Sutton \& Barto (1998, 2018) and Bertsekas \& Tsitsiklis (1996).

A MDP is defined by a set of states, $S$, a set of actions, $A$, a reward
function, $R(s,a)$ over state/action pairs, and a state transition distribution,
$P(s'|s,a)$, where $a$ denotes a chosen action. States and rewards occur
sequentially according to these one-step functions, driven by a series of
actions; the goal is to learn to choose a probabilistic policy over actions,
denoted by $\pi$, that maximizes the value function, $V^\pi(s)$, defined as the
expected cumulative discounted reward:

@@

Here, $\gamma$ is a parameter controlling temporal discounting. The value function
can also be defined recursively as the sum of the immediate reward of the action
chosen in that state, $R(s, a)$, and the value of its successor state $s’$,
averaged over possible actions, $a$, and transitions that would occur if the agent
chose according to $\pi$:

$$ V^\pi(s) = \sum_a \pi(a|s) \left[ R(s,a) + \sum_{s'} P(s'|s,a) \gamma V^\pi (S') \right] $$

The value function under the optimal policy is given by:

@@

$$ = \max_a \left[ R(s,a) + \sum_{s'} P(s'|s,a) \gamma V^\pi (S') \right] $$

Knowledge of the value function can help to guide choices. For instance, we can
define the state-action value function as the value of choosing action $a$ and
following $\pi$ thereafter:

@@

$$ = R(s,a) + \sum_{s'} P(s'|s,a) \gamma V^\pi (S') $$

Then at any state one could choose the action that maximizes $Q^\pi(s,a)$. Formally
this defines a new policy, which is as good or better than the baseline policy
$\pi$; analogously, Eq 2 can be used to define the optimal state-action value
function, the maximization of which selects optimal actions. Note that it is
possible to write a recursive definition for $Q$ in the same manner as Eq 1, and
work directly with the state-action values, rather than deriving them indirectly
from $V$.

\subsection{Robust Control}

To create the symptoms of anxiety, we need a loss of control. This can come in
two forms. First is through the state transition distribution, $P(s'|s,a)$.
Normally, we assume the world is deterministic; in other words, an agent's chosen
action, $a$ leads the agent to the desired successor state, $s'$. In a stochastic
environment, however, an agent may instead transition to an undesired state, $s^*$.

Importantly, in the clinical literature, pathological anxiety is also associated
with a lack of perceived control (Gallagher et al., 2014). Studies suggest that
this belief mediates anxiety (ref?). This can come in whatever form.

Alternately, an agent may choose optimally in the present but may be uncertain of
its choice control in the future. In such a case, we deviate from the optimal
policy in the future. We can adopt a different function. There are a number to
choose from \citep{Garcia2015}. Here we adopt the beta-pessimism criterion
\citep{Gaskett2003}:

$$ V(s') = \beta \max_{a'} Q(s',a') + (1 - \beta) \min_{a'} Q(s',a') $$

where $\beta$ controls the degree of pessimism. When $\beta = 1$, an agent expects
complete control to choose the best action (best-case criterion); when $\beta = 0$,
an agent expects the  worst future case; and when $\beta = 0.5$, an agent expects
randomness. (We are not committed to this particular instantiation; it is simply
a convenient notation. Other parameterizations work equally well, see appendix.)

@@

@@

>>>>

\author{
Samuel Zorowitz \\
Princeton Neuroscience Institute\\
Princeton University\\
Princeton, NJ 08540 \\
\texttt{zorowitz@princeton.edu} \\
\And
Ida Momennejad \\
Columbia University\\
New York, NY 10027 \\
\texttt{ida.m@columbia.edu} \\
\And
Nathaniel Daw \\
Princeton Neuroscience Institute\\
Princeton University\\
Princeton, NJ 08540 \\
\texttt{ndaw@princeton.edu} \\
}

@@

\section{Introduction}

Central to many psychiatric disorders is distortion to representations of the
external world and, correspondingly, pronounced changes to learning and decision
making. As a normative framework, decision theory allows us to describe and pose
quantita­tive questions about optimal choice behavior \citep{DayanDaw2008} and is,
therefore, a critical tool for modeling, understanding, and predicting the psychological
and neurobiological mechanisms underlying psychiatric behavior \citep{maia2011, HuysDawDayan2015}.
Recent investigations employing this framework have provided novel and important
insights into psychiatric symptoms and disorders, including anhedonia in major depression \citep{Rutledge2017},
mood instability in bipolar disorder \citep{EldarNiv2015, EldarDolanNiv2016},
habitual behavior in obsessive compulsive and addictive disorders \citep{gillan2016}, and
hallucinations in schizophrenia \citep{powers2017, corlett2018}.

The family of anxiety disorders (which includes generalized anxiety disorder, social
anxiety disorder, panic disorder, and specific phobia) are characterized by excessive
fear and worry about future negative outcomes, and represent the most common
group of psychiatric disorders \citep{kessler2009}. Though the disorders typically
differ in the types of objects or situations that induce fear or anxiety, all share
a set of core disturbances to cognition and behavior. First, the anxiety disorders
involve pessimistic inference, or a tendency to appraise threatening objects or
situations as more likely and/or more severe than they objectively are. Clinical
anxiety is also associated with second-order conditioning, or learning to fear
neutral cues indirectly associated with threat cues. Finally, anxiety is associated
with sustained avoidance, or the tendency for individuals to flee or avoid threatening
situations. As an example, an individual is mugged in an alleyway. In a pathological
response, this individual may exhibit persistent worry that it will happen again
(pessimistic inference), increased anxiety towards the street or time of day
at which it occurred (second-order conditioning), and increased effort to stay
away from any cue which reminds them of the event (avoidance).

Several studies have attempted to provide some insight into why some of this
may be in anxiety disorders \cite{Maia2010, norbury2018}, though no account has
tied these symptoms together. For example, \citep{Maia2010} provided a simple
account of how simple actor-critic reinforcement learning models could account
for sustained avoidance in rodent shuttle-box experiments \citep{servatius2008}.
As has been discussed elsewhere \citep{Moutoussis2017}, such an account fails to account for why avoidance
is maintained even with non-threatening outcomes and do not account for why exposure
therapy only works for some individuals. Other accounts have attempted to explain
pessimistic inference (cite learning rate), but do not mention where from.

In the remainder of this article, we present a simple decision theoretic account
of how the core symptoms of anxiety may arise and be maintained. Specifically, we
demonstrate that the loss of optimistic inference in the form of reduced expectations
of future reward can yield negatively biased estimates of value in sequential
decision environments. In such an event, pessimistic inference, second-order
conditioning, and chronic avoidance naturally follow.

@@

Before moving forward, we first briefly review the formalism of Markov decision
processes, which provide the framework for our results. For complete treatments,
see \cite{SuttonBarto1998, SuttonBarto2018} and \cite{bertsekas2005}.

A MDP is defined by a set of states, $S$, a set of actions, $A$, a reward
function defined over state/action pairs, $R(s,a)$, and a state transition distribution,
$P(s'|s,a)$, where $a$ denotes a chosen action. In a MDP, states and rewards occur
sequentially according to the one-step transition structure where transitions follow
actions. The goal for an agent in a MDP is to learn to choose a policy, $\pi$, or
mapping of state to action, that maximizes long-run expected reward. Here we define
the value of a state under a policy, $V^\pi(s)$, as the expected cumulative discounted reward:

@@

where $\gamma$ is the temporal discounting parameter, controlling the weighting
of distant future rewards. The value function can be defined recursively as the
sum of the immediate reward following an action chosen in a state, $R(s, a)$, and
the value of its successor state $s’$, averaged over potential future actions, $a$,
and transitions according to the current policy $\pi$:

$$ V^\pi(s) = \sum_a \pi(a|s) \sum_{s',r}p(s',r|s,a) \left[ r + \gamma v(s') \right] $$

Under the optimal policy, the value function is given by:

@@

$$ = \max_a \sum_{s',r}p(s',r|s,a) \left[ r + \gamma v_*(s') \right] $$

Using the formalisms above, the value of one particular action in a state under
the optimal policy is defined as:

@@

$$ = \sum_{s',r}p(s',r|s,a) \left[ r + \gamma v_*(s') \right] $$

Insofar that the value of a state under the optimal policy is defined as the best
action under that state, we can re-express the above equivalently as:

$$ q_*(s,a) = \sum_{s',r}p(s',r|s,a) \left[ r + \gamma \max_{a'} q_*(s',a') \right] $$

Thus in a conventional MDP, the value of states in the environment under the optimal
policy assumes the agent is able to act according to the optimal policy in the future.

\subsection{Robust Control}
In the previous section, we observed that the value of a state and its associated
state-action pairs are a function of the environment transition distribution and
the best future action. In standard MDP problems, it is typically assumed that the
transition distribution is known and accurate, and that the agent is consistently
able to operate under the optimal policy in the future. In this section, we introduce
a formalism deviating from these standard assumptions.

Central to many theories of clinical anxiety is a perceived loss of control (see
discussion for elaboration). In the MDP framework described above, this may come
in one of several forms. First, an agent may perceive the transition structure
of the environment to be less reliable than it actually is. Put another way, an
agent may have a biased expectation that, despite choosing an action $a$ with the
intent of transition to state $s'$, the agent may instead transition to an alternate,
undesired state $s^$. In the case of clinical anxiety, the expectation may be more
dire, that the environment is "adversarial" is likely to transition an agent, not
randomly, but to the worse states.

Second, an agent may doubt its own agentive control. An agent may expect to be
able to choose the best possible action for itself in the present, but may not be
confident it will still be able to decide optimally for itself in the future. In
either event, following the value computations defined above, we should expect
the value estimates for such an agent to diminish. If an agent believes itself
or the environment to be unreliable, then the agent cannot be optimistic about
its future and should instead adopt a different policy.

These sorts of problems have been explored in the safe reinforcement learning
and robust control literature \citep{Garcia2015}. There, we are concerned
with programming agents that identify the optimal policy under pessimistic
assumptions. A number of approaches have been proposed in that literature. Here,
we focus on one particular simple implemented proposed by \cite{Gaskett2003}, namely
*w-pessimism*:

$$ V(s') = w \max_{a'} q(s',a') + (1 - w) \min_{a'} q(s',a') $$

such that the state value under the optimal policy is defined as:

$$ V(s) = \max_a \sum_{s',r}p(s',r|s,a) \left[ r + \gamma \left( w \max_{a'} q(s',a') + (1 - w) \min_{a'} q(s',a') \right) \right] $$

In this formalism, $w$ controls the degree of pessimism. When $\w = 1$, an agent
is maximally optimistic and fully expects to act according to the best possible actions
in the future. When $w = 0$, the agent is maximally pessimistic and fully expects
to act according to the worst possible actions in the future. When $w = 0.5$,
the weighs the best and worst possible future actions equally.

We note that we are not committed to this particular implementation of a pessimistic
agent. As mentioned above, an alternative parameterization would have been to
program an agent with a belief in an adversarial environment, preferentially
transitioning it to worse states. Insofar that these two terms are multiplicative
in value calculations, either implementation should yield similar results. Similarly,
the w-pessimism function could easily be replaced with other functions (e.g. softmax
with negative inverse temperature). This formalism is simply convenient and is
useful for demonstrating the points we demonstrate below. It is not a substantive
claim about anxiety.

@@

In self-report,
individuals with anxiety rate hypothetical future negative outcomes as
being more likely to occur and worse to experience than non-anxious counterparts
\citep{ButlerMathews1983, ButlerMathews1987, Foa1996, MacLeod1996, MacLeod1997,
Luten1997, Stober1997, Borkovec1999, Maner2006, Mitte2007, Miranda2007}. In the
laboratory, individuals with anxiety demonstrate excessive fear in fear conditioning
paradigms. For example, anxiety is correlated with exaggerated fear towards
non-threatening cues during conditioning and extinguished cues during extinction
\citep{lissek2005, Duits2015}.

@@

Avoidance is often chosen even at the expense of potential reward (approch-avoidance
conflict), and is frequently a major impediment to recovery \citep{Arnaudova2017}.
